# Data Visualisation with Grammar of Graphics

## Plotting with R

- with base R

- ggplot2 package: `ggplot()` function


## Grammar of graphics

The Grammar of Graphics is a structured way of thinking about and building data visualizations

**Key Components of the Grammar of Graphics**

Data: the dataset you want to visualize.

Aesthetics (aes): how variables are mapped to visual properties (x, y, color, size, shape).

Geometries (geoms): the type of plot you draw (points, lines, bars, boxplots).

Scales:  control how data values map to aesthetics (e.g., continuous vs. categorical colors, log scales).



Coordinate system:  the space in which the data is drawn (Cartesian, polar, map projections).

Facets: splitting data into subplots for comparison.


Statistical transformations (stats):  summaries or transformations (e.g., binning in histograms, smoothing in regression lines).

Themes:  non-data elements like fonts, backgrounds, grid lines.

## Ingredients for plotting

1. Data


```{r}
data("mtcars")
head(mtcars)
```

2. Canvas to draw the plot

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
ggplot()
```


## Data + Aesthetics (aes)


```{r, comment=NA, fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg))
```


```{r, comment=NA, fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg,
    col=as_factor(cyl)))
```

## Data + Aesthetics (aes) + Geometries (geoms)


```{r, comment=NA,   fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg)) +
  geom_point()
```


```{r, comment=NA,   fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg)) +
  geom_point()
```


```{r, comment=NA,  fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg)) +
  geom_smooth()
```


```{r, comment=NA,  fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg)) +
  geom_density_2d()
```

```{r, comment=NA,  fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars,
aes(x=wt, 
    y= mpg)) +
  geom_density_2d() + 
  geom_point()
```


## Add  `scale` layer


```{r, comment=NA,  fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point() +
  scale_color_brewer(palette = "Set1")

```


```{r, comment=NA,  fig.height=5.5, warning=FALSE, message=FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point() +
  scale_color_manual(values = c("red", "blue", "green"))
```


## Add `coord` layer

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(xlim = c(2, 5), ylim = c(10, 35))

```


```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  coord_fixed(ratio = 3/10)

```

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  coord_flip()
```

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  coord_polar()
```

## Add `facet` layer


```{r}
ggplot(mtcars, aes(x = wt, y = mpg, col = as_factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE) +
  facet_wrap(~cyl)
```

## Add `stat` layer

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  stat_summary(fun = "mean", geom = "point", size = 5, shape = 18, col="red", alpha=0.5) +
  scale_color_brewer(palette = "Set1")

```

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  stat_summary(fun = "mean", geom = "point", size = 5, shape = 18, col="red", alpha=0.5) +
    geom_point() +
  scale_color_brewer(palette = "Set1")

```

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  stat_summary(fun = "mean", geom = "point", size = 5, shape = 18, col="green", alpha=0.5) +
  scale_color_brewer(palette = "Set1")

```


## theme layer

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  stat_summary(fun = "mean", geom = "point", size = 5, shape = 18, col="red", alpha=0.5) +
    geom_point() +
  scale_color_brewer(palette = "Set1") +
  theme_bw()

```

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  stat_summary(fun = "mean", geom = "point", size = 5, shape = 18, col="red", alpha=0.5) +
    geom_point() +
  scale_color_brewer(palette = "Set1") +
  theme_dark()

```

## Some examples with gapminder dataset

```{r}
library(gapminder)
data(gapminder)
```

 Visualize the relationship between life expectancy, GDP per capita and continent in 2007.


```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
gapminder2007 <- gapminder %>%
  filter(year==2007)
ggplot(gapminder2007,
aes(x=lifeExp, y= gdpPercap,
    col=continent))+ geom_point()+
  theme(legend.position = "bottom") + 
labs(title="Relationship between life expectancy and GPD per capita by continent - 2007",
        x ="life expectancy at birth, in years",
     y = "GDP per capita (US$, inflation-adjusted)")
```
]



### Add a vertical line



```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
gapminder2007 <- gapminder %>%
  filter(year == 2007)
  
ggplot(gapminder2007,
aes(x = lifeExp, y = gdpPercap, col=continent)) +
    geom_point() + 
    geom_vline(xintercept = 70)
```


### Add a horizontal line


```{r, comment=NA,fig.height=6, warning=FALSE, message=FALSE}
gapminder2007 <- gapminder %>%
  filter(year == 2007)
  
ggplot(gapminder2007,
aes(x = lifeExp, y = gdpPercap, col=continent)) +
    geom_point() + 
    geom_hline(yintercept = 20000)
```


### Add a diagonal line


```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
gapminder2007 <- gapminder %>%
  filter(year == 2007)
  
ggplot(gapminder2007,
aes(x = lifeExp, y = gdpPercap, col=continent)) +
    geom_point() + 
    geom_abline(intercept = 20, slope=200)
```

## All geoms in ggplot2

```{r, echo=FALSE, message=FALSE, comment=NA}
ls(pattern = '^geom_', env = as.environment('package:ggplot2'))

```


## geom_boxplot


```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
geom_boxplot()
```

```{r, comment=NA,   fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent, color=continent))+
geom_boxplot()
```

```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent, fill=continent))+
geom_boxplot()
```
]


```{r, comment=NA,   fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
geom_boxplot(fill="forestgreen")
```
]


```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
geom_boxplot(fill="forestgreen", alpha=0.5)
```
]


## geom_point



```{r, comment=NA,  echo=FALSE, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
  geom_point()
```



##  geom_jitter


```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
  geom_jitter()
```



##  geom_jitter + geom_boxplot


```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
  geom_jitter() +
  geom_boxplot()
```

```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
  geom_jitter() +
  geom_boxplot(alpha=0.5)
```



```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent))+
  geom_boxplot() + 
  geom_jitter() 
```


```{r}
ggplot(gapminder2007, aes(x=lifeExp, y=continent, fill=continent))+
  geom_boxplot() + 
  geom_jitter(aes(col=continent)) 
```

**geom_jitter + geom_boxplot (outlier.shape = NA)**


```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent, fill=continent))+
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(aes(col=continent)) 
```


.right-plot[
```{r, comment=NA,  echo=FALSE, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, aes(x=lifeExp, y=continent, fill=continent, col=continent))+
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(aes(col=continent)) 
```

Write the code to obtain the following plot.


```{r, comment=NA,  echo=FALSE, fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, 
aes(x=lifeExp, col=continent, fill=continent))+
  geom_density(alpha=0.5) 
```

##  geom_histogram


```{r, comment=NA,  fig.height=6, warning=FALSE, message=FALSE}
ggplot(gapminder2007, 
aes(x=lifeExp))+
  geom_histogram() 
```


## geom_bar (stat="identity")


```{r, comment=NA}
cut.percent <- data.frame(cut=c("Fair", "Good", "Very Good", "Premium", 
                                "Ideal"), percent=c(3, 9, 22.4, 25.6, 40))
cut.percent
    
```



```{r, comment=NA,  fig.height=3, warning=FALSE, message=FALSE}
ggplot(data=cut.percent, aes(x=cut, y=percent)) +
  geom_bar(stat="identity")
```


## geom_col



```{r, comment=NA}
cut.percent <- data.frame(cut=c("Fair", "Good", "Very Good", "Premium", 
                                "Ideal"), percent=c(3, 9, 22.4, 25.6, 40))
cut.percent
    
```



```{r, comment=NA,  fig.height=3, warning=FALSE, message=FALSE}
ggplot(data=cut.percent, aes(x=cut, y=percent)) +
  geom_col()
```


# geom_line



```{r, comment=NA, fig.height=6, warning=FALSE, message=FALSE}
gapminder %>%
filter(country == "India") %>%
ggplot(aes(x = year, y = gdpPercap)) +
geom_line() 
```

Your turn

Write the code to obtain the following plot.

```{r, comment=NA,  echo=FALSE, fig.height=6, warning=FALSE, message=FALSE}
gapminder %>%
filter(country == "India") %>%
ggplot(aes(x = year, y = gdpPercap)) +
geom_line() + 
  geom_point() + 
  labs(title="Time series plot of GDP per capita",
       x="Year", 
       y="GDP per capita (USD, inflation-adjusted)")
```



## Data Wrangling + Data Visualization

```{r, comment=NA}
avglifeExp <- gapminder %>%
  group_by(continent, year) %>%
  summarise(meanlifeExp=mean(lifeExp))
avglifeExp
```



```{r, comment=NA, message=FALSE, fig.height=5}
ggplot(avglifeExp, aes(x=year, y=meanlifeExp, col=continent))+
  geom_line() + geom_point()
```

## Exercise

Write an R code to reproduce the plot below. 


```{r, comment=NA, message=FALSE, fig.height=5, echo=FALSE}
gapminder %>%
  filter(year %in% c(1952, 1957, 1962, 1967, 1972, 1977)) %>%
  filter(continent %in% c("Asia", "Americas")) %>%
  ggplot(aes(y=lifeExp, x=gdpPercap, color=continent)) + 
  geom_point() + 
  facet_wrap(~year, ncol=3)+
  labs(title="Life Expectancy vs GDP - America and Asia",
       y = "Life Expectancy",
       x = "GDP per capita")
```


Write an R code to reproduce the plot below. 


```{r, comment=NA, message=FALSE, fig.height=5, echo=FALSE}
gapminder %>%
  filter(year %in% c(1952, 1957, 1962, 1967, 1972, 1977)) %>%
  filter(continent %in% c("Asia", "Americas")) %>%
  ggplot(aes(y=log(lifeExp), x=log(gdpPercap), color=continent)) + 
  geom_point() + 
  facet_wrap(~year, ncol=3)+
  labs(title="Life Expectancy vs GDP - America and Asia",
       y = "log(Life Expectancy)",
       x = "log(GDP per capita)")
```



Write an R code to reproduce the plot below. 


```{r, comment=NA, message=FALSE, fig.height=5, echo=FALSE}
ggplot(gapminder_unfiltered, aes(gdpPercap, lifeExp, color = year)) + 
geom_point() +
facet_wrap(~ continent)
```


Write an R code to reproduce the plot below. 


```{r, comment=NA, message=FALSE, fig.height=5, echo=FALSE}
ggplot(gapminder, aes(y=lifeExp,  x=year)) + 
geom_smooth() +
  geom_point() + 
facet_wrap(~ continent)
```



