# Regression Analysis

## Introduction

## Simple Linear Regression

## Multiple Linear Regression

## Regression with Dummy Variables

## Exercise

1. To understand the effectiveness of marketing investments, a manager of a company wants to analyze the relationship between marketing expenditures (USD) and the number of sales. By fitting a simple linear regression model with marketing spend as the independent variable and sales as the dependent variable. Interpret the outputs.

```{r, echo=FALSE}
library(kableExtra)
set.seed(2024727)
error <- rnorm(20, sd=2)
me <- c(100, 115, 120, 132, 144, 154, 158, 160, 170, 180, 181, 190, 200, 210, 220, 230, 240, 250, 260, 270)
sales <- round(5*me + 20 + error, 1)
df <- data.frame(me = me, sales=sales)
df |> knitr::kable( col.names = c("Marketing expenditures (USD)",
                           "Number of Sales")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) |>
  column_spec(1:2, text_align = "center")
```


2. To understand the effectiveness of marketing investments, a manager of a company wants to analyze the relationship between marketing expenditures (USD) and the number of sales in two locations. By fitting a multiplelinear regression model with marketing spend and location as the independent variables and sales as the dependent variable. Interpret the outputs.

```{r, echo=FALSE}
set.seed(2024727)
error <- rnorm(20, sd=2)
me1 <- c(100, 115, 120, 132, 144, 154, 158, 160, 170, 180, 181, 190, 200, 210, 220, 230, 240, 250, 260, 270)
me2 <- c(101, 115, 122, 132, 145, 154, 158, 160, 170, 180, 181, 191, 200, 210, 220, 230, 240, 250, 260, 270) + 2
sales1 <- round(5*me1 + 20 + error, 1)
sales2 <- round(5*me1 + 20 + error, 1)
sales <- c(sales1, sales2)
location <- c(rep("A", 20), rep("B", 20))
df <- data.frame(me = me, sales=sales, location=location)
df |> knitr::kable( col.names = c("Marketing expenditures (USD)",
                           "Number of Sales", "Location")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) |>
  column_spec(1:3, text_align = "center")
```